High geomagnetic field intensity in the late Mesoproterozoic recorded by Midcontinent Rift anorthosite xenoliths

### 2021-10-12 manuscript/Manuscript.tex

However, the updated paleointensity records from the Midcontinent Rift together with previous results from other Mesoproterozoic rocks suggest that the a dramatic increase in geomagnetic dipole moment similar in scale to that modeled for after the onset of inner core nucleation may not be unique to the Edicaran-Cambrian boundary before 500 Ma (Figure \ref{fig:PINT_compilation}). The \textit{ca.} 1140 Ma Abitibi dikes \cite{Macouin2003a} record an average virtual dipole moment of $\sim$13 ZAm$^2$, similar to the low values of $\sim$7 ZAm$^2$ obtained from the single silicate crystals of the \textit{ca.} 565 Ma Sept-\hat{I}les intrusions \cite{Bono2019a}. However, the relatively high virtual dipole moments of $\sim$72 ZAm$^2$ from the \textit{ca.} 1105 Ma Osler Volcanic Group \cite{Sprain2018a} and $\sim$82 ZAm$^2$ from the \textit{ca.} 1092 Ma Beaver River anorthosite xenoliths (Figure \ref{fig:PINT_compilation} are about six times the values from the \textit{ca.} 1140 Ma Abitibi dikes. In addition, the increase in absolute dipole moment estimates recorded by the Midcontinent Rift rocks are much higher than that in the Ediacaran (Figure \ref{fig:PINT_compilation}).

Furthermore, the high geomagnetic field estimates from the Midcontinent Rift volcanics and intrusives are not the only suite of rocks that bear records at odds with the hypothesis of a progressive decay trend during the Proterozoic \cite{Biggin2009a, Bono2019a}. The $\sim$60 ZAm$^2$ dipole moment estimate from the \textit{ca.} 755 Ma Mundine Well dikes are discordant with the long-term decrease trend \cite{Bono2019a} and is about three times the estimates from the \textit{ca.} 720 Ma dolerite dikes of the Franklin Large Igneous Province \cite{Lloyd2021a}. Similarly, an apparent rapid decrease in geomagnetic dipole moment from \textit{ca.} 1092 Ma to \textit{ca.} 1070 Ma can be evidenced by the contrasting high values from the Midcontinent Rift anorthosite xenoliths and the low values from the Mundine Well dikes (Figure \ref{PINT_Compilation}). However, a sparse record in a short time period may not sufficiently average out paleosecular variations. Given that a five-fold fluctuation of the geomagnetic dipole moment exist in the paleointensity records for the past 10 million years (5 and 95 percentiles), when we know a modern-day-like geodynamo likely powered by both thermal and compositional convection existed, we cannot conclude on whether an apparent rise or decay within a narrow time period ($\sim$35 Ma) during the Proterozoic is associated with transitions in the core fluid regime or is a result of short-term variations of the geomagnetic field that may be common through much of the evolution of Earth's core. In addition, the observations of geomagnetic dipole moment fluctuations in more recent time scales may suggest that evaluations of long-term evolution of the Precambrian paleointensity records should be accompanied by development of higher-resolution data in more short-term scales.

Overall, the existing absolute paleointensity records in the PINT database remains sparse in terms of spatial and temporal coverage as compared to more geologically recent times. (Figure \ref{PINT_compilation}). More data are needed in order to better characterize the evolution of Earth's dynamo and thermal evolution.

@@

>>>>

However, the updated paleointensity records from the Midcontinent Rift together with previous results from other Mesoproterozoic rocks suggest that the a dramatic increase in geomagnetic dipole moment similar in scale to that modeled for after the onset of inner core nucleation may not be unique to the Edicaran-Cambrian boundary before 500 Ma (Figure \ref{fig:PINT_compilation}). The \textit{ca.} 1140 Ma Abitibi dikes \cite{Macouin2003a} record an average virtual dipole moment of $\sim$13 ZAm$^2$, similar to the low values of $\sim$7 ZAm$^2$ obtained from the single silicate crystals of the \textit{ca.} 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a}. However, the relatively high virtual dipole moments of $\sim$72 ZAm$^2$ from the \textit{ca.} 1105 Ma Osler Volcanic Group \cite{Sprain2018a} and $\sim$82 ZAm$^2$ from the \textit{ca.} 1092 Ma Beaver River anorthosite xenoliths (Figure \ref{fig:PINT_compilation} are about six times the values from the \textit{ca.} 1140 Ma Abitibi dikes. In addition, the increase in absolute dipole moment estimates recorded by the Midcontinent Rift rocks are much higher than that in the Ediacaran (Figure \ref{fig:PINT_compilation}).

Furthermore, the high geomagnetic field estimates from the Midcontinent Rift volcanics and intrusives are not the only suite of rocks that bear records at odds with the hypothesis of a progressive decay trend during the Proterozoic \cite{Biggin2009a, Bono2019a}. The $\sim$60 ZAm$^2$ dipole moment estimate from the \textit{ca.} 755 Ma Mundine Well dikes are discordant with the long-term decrease trend \cite{Bono2019a} and is about three times the estimates from the \textit{ca.} 720 Ma dolerite dikes of the Franklin Large Igneous Province \cite{Lloyd2021a}. Similarly, an apparent rapid decrease in geomagnetic dipole moment from \textit{ca.} 1092 Ma to \textit{ca.} 1070 Ma can be evidenced by the contrasting high values from the Midcontinent Rift anorthosite xenoliths and the low values from the Mundine Well dikes (Figure \ref{fig:PINT_compilation}). However, a sparse record in a short time period may not sufficiently average out paleosecular variations. Given that a five-fold fluctuation of the geomagnetic dipole moment exist in the paleointensity records for the past 10 million years (5 and 95 percentiles), when we know a modern-day-like geodynamo likely powered by both thermal and compositional convection existed, we cannot conclude on whether an apparent rise or decay within a narrow time period ($\sim$35 Ma) during the Proterozoic is associated with transitions in the core fluid regime or is a result of short-term variations of the geomagnetic field that may be common through much of the evolution of Earth's core. In addition, the observations of geomagnetic dipole moment fluctuations in more recent time scales may suggest that evaluations of long-term evolution of the Precambrian paleointensity records should be accompanied by development of higher-resolution data in more short-term scales.

Overall, the existing absolute paleointensity records in the PINT database remains sparse in terms of spatial and temporal coverage as compared to more geologically recent times. (Figure \ref{fig:PINT_compilation}). More data are needed in order to better characterize the evolution of Earth's dynamo and thermal evolution.

@@

% question to ask about the paleointensity database:
